---
title: "F2 adaptations to dominant markers"
author: "Cristiane Hayumi Taniguti - Statistical Genetics Lab - Department of Genetics - Luiz de Queiroz College of Agriculture - University of São Paulo"
output: pdf_document
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)

knitr::opts_chunk$set(echo=TRUE)
```

OneMap until version 2.2.0 does not adequately deal with dominant markers in F2 populations. The modification was to apply phase estimation to F2 populations using the same background algorithms used for outcrossing populations. This modification was needed to estimate the correct phase between dominant and codominant markers. The modification also allows  the estimation of the phase between the heterozygotes progeny in codominant markers. In this new version, we want to provide to users progeny haplotypes, and for that, it's essential to distinguish phase also in F2 intercross. 

**Warning**: Because of all the modifications, users can find differences in estimations for this type of population between this and older versions of OneMap.

Here we perform some simulations to show these differences and the need for updates. Two scenarios will be simulated. One with only codominant markers and other with codominant and dominant markers. Three metacentric chromosomes compose the maps with 100 cM and population size of 200 individuals. 

* Scenario 1: 150 codominant (A.H.B) markers, 50 per chromosome and no missing data
* Scenario 2: 50 codominant (A.H.B), 50 D.B markers, and 50 C.A markers, 50 markers per chromosome and no missing data

Both scenarios will be evaluated by the old and new approaches.

# Packages

```{r,colFmt("red")}
library(onemap)
```

# Scenario 1

## Simulation

```{r, eval=FALSE}
run_pedsim(chromosome = c("Chr1", "Chr02", "Chr03"), n.marker = c(50,50,50), 
           tot.size.cm = c(100,100,100), centromere = c(50, 50, 50),
           n.ind = 200, mk.types = c("A.H.B"), 
           n.types = c(150), pop = "F2", 
           path.pedsim = "/home/cristiane/Programs/PedigreeSim/",
           name.mapfile = "mapfile.map", name.founderfile="founderfile.gen",
           name.chromfile="sim.chrom", name.parfile="sim.par",
           name.out="sim_cod_F2")
```

```{r, eval=FALSE}
pedsim2raw(cross="f2 intercross", genofile = "sim_cod_F2_genotypes.dat", 
           parent1 = "P1", parent2 = "P2", f1 = "F1", 
           out.file = "sim_cod_F2.raw", miss.perc = 0)
```

## New approach

```{r, eval=FALSE}
dataset <- read_onemap("sim_cod_F2.raw")

twopts <- rf_2pts(dataset)
seq1 <- make_seq(twopts, "all")
lod_sug <- suggest_lod(dataset)
lgs <- group(seq1, LOD= lod_sug) # Groups is correct

lg1 <- make_seq(lgs,1)
# Distance estimation
time <- system.time(map1 <- map(lg1)) 
save.obj <- list(map1, time)
save(save.obj, file = "save.obj4.RData")
```

```{r, echo=FALSE}
load("save.obj4.RData")
save.obj[[1]]

cat("Time spent")
save.obj[[2]]
```

```{r, eval=FALSE}
# Ordering
ug1 <- ug(lg1)
rcd1 <- rcd(lg1)
seriation1 <- seriation(lg1)
record1 <- record(lg1)
mds1 <- mds_onemap(lg1)
order1 <- order_seq(lg1)
order1 <- make_seq(order1, "force")

p_ug_cod1 <- rf_graph_table(ug1)
p_rcd_cod1 <- rf_graph_table(rcd1)
p_ser_cod1 <- rf_graph_table(seriation1)
p_rec_cod1 <- rf_graph_table(record1)
p_mds_cod1 <- rf_graph_table(mds1)
p_map_cod1 <- rf_graph_table(map1)
p_order_cod1 <- rf_graph_table(order1)
```

## Old approach

```{r, eval=FALSE}
dataset <- read_onemap(inputfile = "sim_cod_F2.raw")

twopts <- rf_2pts(dataset)
seq1 <- make_seq(twopts, "all")
lod_sug <- suggest_lod(dataset)
lgs <- group(seq1, LOD= lod_sug) # Group is correct

lg1 <- make_seq(lgs,1)
# Distance estimation
time <- system.time(map1 <- map(lg1))

save.obj <- list(map1, time)
save(save.obj, file = "save.obj3.RData")
```

```{r, echo=FALSE, eval=FALSE}
# Can not be printed, new version new sequence object
load("save.obj3.RData")
save.obj[[1]]

cat("Time spent")
save.obj[[2]]
```

```
Printing map:

Markers           Position           

 1 M001               0.00           
 2 M002               1.76           
 3 M003               4.29           
 4 M004               6.06           
 5 M005               8.60           
 6 M006              12.44           
 7 M007              13.19           
 8 M008              14.70           
 9 M009              16.73           
10 M010              18.24           
11 M011              20.26           
12 M012              22.55           
13 M013              24.32           
14 M014              26.08           
15 M015              27.60           
16 M016              29.36           
17 M017              31.39           
18 M018              33.41           
19 M019              34.92           
20 M020              36.68           
21 M021              38.19           
22 M022              40.73           
23 M023              43.00           
24 M024              45.02           
25 M025              48.33           
26 M026              49.34           
27 M027              51.10           
28 M028              52.62           
29 M029              54.13           
30 M030              55.89           
31 M031              57.40           
32 M032              60.45           
33 M033              61.45           
34 M034              64.76           
35 M035              67.80           
36 M036              71.37           
37 M037              72.63           
38 M038              74.14           
39 M039              76.93           
40 M040              78.94           
41 M041              81.48           
42 M042              82.73           
43 M043              85.01           
44 M044              85.76           
45 M045              87.02           
46 M046              88.78           
47 M047              90.55           
48 M048              93.08           
49 M049              93.83           
50 M050              95.85           

 50 markers            log-likelihood: -1920.484 

Time spent

user  system elapsed 
0.016   0.000   0.013 

```

```{r, eval=FALSE}
# Ordering
ug1 <- ug(lg1)
rcd1 <- rcd(lg1)
seriation1 <- seriation(lg1)
record1 <- record(lg1)
mds1 <- mds_onemap(lg1)
order1 <- order_seq(lg1)
order1 <- make_seq(order1, "force")

p_ug_cod_old1 <- rf_graph_table(ug1)
p_rcd_cod_old1 <- rf_graph_table(rcd1)
p_ser_cod_old1 <- rf_graph_table(seriation1)
p_rec_cod_old1 <- rf_graph_table(record1)
p_mds_cod_old1 <- rf_graph_table(mds1)
p_map_cod_old1 <- rf_graph_table(map1)
p_order_cod_old1 <- rf_graph_table(order1)
```

# Scenario 2

## Simulation

```{r, eval=FALSE}
run_pedsim(chromosome = c("Chr1", "Chr02", "Chr03"), n.marker = c(50,50,50), 
           tot.size.cm = c(100,100,100), centromere = c(50, 50, 50),
           n.ind = 200, mk.types = c("A.H.B", "C.A", "D.B"), 
           n.types = c(50, 50, 50), pop = "F2", 
           path.pedsim = "/home/cristiane/Programs/PedigreeSim/",
           name.mapfile = "mapfile.map", name.founderfile="founderfile.gen",
           name.chromfile="sim.chrom", name.parfile="sim.par",
           name.out="sim_F2")
```

```{r, eval=FALSE}
pedsim2raw(cross="f2 intercross", genofile = "sim_F2_genotypes.dat", 
           parent1 = "P1", parent2 = "P2", f1 = "F1", 
           out.file = "sim_F2.raw", miss.perc = 0)
```

## New approach

```{r, eval=FALSE}
dataset <- read_onemap(inputfile = "sim_F2.raw")

twopts <- rf_2pts(dataset)
seq1 <- make_seq(twopts, "all")
lod_sug <- suggest_lod(dataset)
lgs <- group(seq1, LOD= lod_sug) # Do not group correctly

lg1 <- make_seq(twopts,1:50)
# Distance estimation
time <- system.time(map1 <- map(lg1)) 
save.obj <- list(map1, time)

save(save.obj, file = "save.obj2.RData")
```

```{r, echo=FALSE}
load("save.obj2.RData")
save.obj[[1]]

cat("Time spent")
save.obj[[2]]
```

```{r, eval=FALSE}
# Ordering
ug1 <- ug(lg1)
rcd1 <- rcd(lg1)
#seriation1 <- seriation(lg1) Error: There are 
# too many ties in the ordering process - please, 
# consider using another ordering algorithm.
record1 <- record(lg1)
mds1 <- mds_onemap(lg1)
order1 <- order_seq(lg1)
order1 <- make_seq(order1, "force")

p_ug1 <- rf_graph_table(ug1)
p_rcd1 <- rf_graph_table(rcd1)
#p_ser1 <- rf_graph_table(seriation1)
p_rec1 <- rf_graph_table(record1)
p_mds1 <- rf_graph_table(mds1)
p_map1 <- rf_graph_table(map1)
p_order1 <- rf_graph_table(order1)
save.image(file = "new_app.RData")
```

## Old approach

```{r, eval=FALSE}
dataset <- read_onemap(inputfile = "sim_F2.raw")

twopts <- rf_2pts(dataset)
seq1 <- make_seq(twopts, "all")
lod_sug <- suggest_lod(dataset)
lgs <- group(seq1, LOD= lod_sug) # Do not group correctly

lg1 <- make_seq(twopts,1:50)
# Distance estimation
time <- system.time(map1 <- map(lg1)) # Time spent 2.150 sec

save.obj <- list(map1, time)
save(save.obj, file = "save.obj1.RData")
```

```{r, echo=FALSE, eval=FALSE}
# New version new sequence object, it can not be printed
load("save.obj1.RData")

save.obj[[1]]

cat("Time spent")
save.obj[[2]]
```

```
Printing map:

Markers           Position           

 1 M001               0.00           
 2 M002             402.95           
 3 M003             805.90           
 4 M004             809.75           
 5 M005             810.92           
 6 M006            1213.88           
 7 M007            1616.83           
 8 M008            2019.78           
 9 M009            2422.73           
10 M010            2825.69           
11 M011            3228.64           
12 M012            3230.64           
13 M013            3231.55           
14 M014            3235.24           
15 M015            3236.92           
16 M016            3239.69           
17 M017            3243.06           
18 M018            3244.23           
19 M019            3247.88           
20 M020            3249.37           
21 M021            3251.74           
22 M022            3654.69           
23 M023            4057.64           
24 M024            4460.59           
25 M025            4863.55           
26 M026            4867.19           
27 M027            4868.85           
28 M028            4868.90           
29 M029            5271.85           
30 M030            5274.38           
31 M031            5677.33           
32 M032            6080.28           
33 M033            6082.05           
34 M034            6085.09           
35 M035            6087.37           
36 M036            6088.87           
37 M037            6491.83           
38 M038            6493.81           
39 M039            6496.31           
40 M040            6899.26           
41 M041            7302.22           
42 M042            7302.22           
43 M043            7303.68           
44 M044            7308.27           
45 M045            7311.35           
46 M046            7714.31           
47 M047            8117.26           
48 M048            8119.71           
49 M049            8121.23           
50 M050            8524.18           

 50 markers            log-likelihood: -4829.173 

Time spent

user  system elapsed 
1.780   0.000   1.785
```

```{r, eval=FALSE}
# Ordering
ug1 <- ug(lg1)
rcd1 <- rcd(lg1)
#seriation1 <- seriation(lg1) Error: There are 
# too many ties in the ordering process - please, 
# consider using another ordering algorithm.
record1 <- record(lg1)
mds1 <- mds_onemap(lg1)
order1 <- order_seq(lg1)
order1 <- make_seq(order1, "force")

p_ug_old1 <- rf_graph_table(ug1)
p_rcd_old1 <- rf_graph_table(rcd1)
#p_ser_old1 <- rf_graph_table(seriation1)
p_rec_old1 <- rf_graph_table(record1)
p_mds_old1 <- rf_graph_table(mds1)
p_map_old1 <- rf_graph_table(map1)
p_order_old1 <- rf_graph_table(order1)
save.image(file = "old_app.RData")
```

# Ordering comparision


```{r, message=FALSE, warning=FALSE}
library(ggpubr)
load("old_app.RData")
load("new_app.RData")

p <- ggarrange(p_ug_cod1, p_ug_cod_old1, p_ug1, 
               p_ug_old1 , common.legend = TRUE,
          labels = c("ug sce1 new", "ug sce1 old", 
                     "ug sce2 new", "ug sce2 old"), 
          vjust = 0.3,
          hjust= -1,
          ncol=2, nrow=2)
p

p <- ggarrange(p_rcd_cod1, p_rcd_cod_old1, p_rcd1, 
               p_rcd_old1 , common.legend = TRUE,
          labels = c("rcd sce1 new", "rcd sce1 old", 
                     "rcd sce2 new", "rcd sce2 old"), 
          vjust = 0.3,
          hjust= -1,
          ncol=2, nrow=2)
p

p <- ggarrange(p_rec_cod1, p_rec_cod_old1, p_rec1, 
               p_rec_old1 , common.legend = TRUE,
          labels = c("rec sce1 new", "rec sce1 old", 
                     "rec sce2 new", "rec sce2 old"), 
          vjust = 0.3,
          hjust= -1,
          ncol=2, nrow=2)
p

p <- ggarrange(p_order_cod1, p_order_cod_old1, p_order1, 
               p_order_old1 , common.legend = TRUE,
          labels = c("order sce1 new", "order sce1 old", 
                     "order sce2 new", "order sce2 old"), 
          vjust = 0.3,
          hjust= -1,
          ncol=2, nrow=2)
p

p <- ggarrange(p_mds_cod1, p_mds_cod_old1, p_mds1, 
               p_mds_old1 , common.legend = TRUE,
          labels = c("mds sce1 new", "mds sce1 old", 
                     "mds sce2 new", "mds sce2 old"), 
          vjust = 0.3,
          hjust= -1,
          ncol=2, nrow=2)
p

```

# Conclusions

With only one repetition of the simulation, we can already see that the modification improves the distance estimation for dominant markers. Also, give more information for the group and ordering algorithms. The OneMap group function does not work properly for these cases; a better approach needs to be implemented. We can also see that ug and MDS algorithms built a better order compared to other ordering algorithms.


# References

Margarido, G. R. A., Souza, A. P., & Garcia, A. A. F. (2007). OneMap: software for genetic mapping in outcrossing species. Hereditas, 144(3), 78–79. https://doi.org/10.1111/j.2007.0018-0661.02000.x
